---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
filename <- "activity.csv"
if(!file.exists(filename)) {
        unzip("activity.zip")
        }
data <- read.csv(filename)

data$date <- as.Date(data$date)
data$hour <- data$interval %/% 100
data$minute <- data$interval %% 100
library(lubridate)
data$DateTime <- ymd_hm(paste0(data$date, " ", data$hour, ":", data$minute))
data$Time <- as.factor(format(data$DateTime, "%H%M"))
```

## What is mean total number of steps taken per day?
```{r echo=TRUE}
steps <- xtabs(steps ~ date, data)

hist(steps)

smean <- mean(steps)
smedian <- median(steps)
```
The mean and median of the total number of steps taken per day are `r as.integer(smean)` and `r as.integer(smedian)`, respectively.

## What is the average daily activity pattern?
```{r echo=TRUE}
library(lattice)
mean_steps <- aggregate(steps ~ Time, data, mean)
End <- length(mean_steps$Time)
Xaxis <- seq(1, End, End/6)
Ylab <- "Number of steps"
xyplot(steps ~ Time, data = mean_steps, type = "l",
       scales = list(x = list(at = Xaxis)), ylab = Ylab)

max_steps <- mean_steps[mean_steps$steps == max(mean_steps$steps), 1]
```
`r max_steps` contains the maximum number of steps on average across all the days in the dataset.

## Imputing missing values
```{r echo=TRUE}
NAs <- sum(apply(data, 1, anyNA))
```
The total number of missing values in the dataset is `r NAs`.

The mean for the respective 5-minute interval is used for filling in all of the missing values in the dataset.
```{r echo=TRUE}
data_fill_na <- data
suppressWarnings(
        for (i in seq_along(data_fill_na$steps)) {
                if(is.na(data_fill_na$steps[i])) {
                        data_fill_na$steps[i] <- mean_steps[mean_steps$Time
                                                    == data_fill_na$Time, 2]
                }
        }
        )

steps_noNA <- xtabs(steps ~ date, data_fill_na)
hist(steps)
smean_noNA <- mean(steps_noNA)
if (smean_noNA == smean) {
        mean_diff <- "the same as"
} else if (smean_noNA > smean) {
        mean_diff <- "larger than"
} else {
        mean_diff <- "smaller than"
}

smedian_noNA <- median(steps_noNA)
if (smedian_noNA == smedian) {
        median_diff <- "the same as"
} else if (smedian_noNA > smedian) {
        median_diff <- "larger than"
} else {
        median_diff <- "smaller than"
}

if(mean_diff == median_diff) {
        also <- "also "
} else {
        also <- ""
}
```
The mean total number of steps taken per day is `r as.integer(smean_noNA)` that is `r mean_diff` the estimates from the first part of the assignment,
whereas the median total number of steps taken per day is `r as.integer(smedian_noNA)` that is `r also``r median_diff` the estimates from the first part of the assignment.

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE}
data_fill_na$Day <- weekdays(data_fill_na$DateTime)
for (i in seq_along(data_fill_na$Day)) {
        if ((data_fill_na$Day[i] == "Saturday") |
                    (data_fill_na$Day[i] == "Sunday")) {
                data_fill_na$Day[i] <- "weekend"
        } else {
                data_fill_na$Day[i] <- "weekday"
        }
}

mean_steps_days <- aggregate(steps ~ Time + Day, data_fill_na, mean)

xyplot(steps ~ Time | Day, data = mean_steps_days, layout = c(1,2), type = "l",
       scales = list(x = list(at = Xaxis)), ylab = Ylab)
```
